
```{r, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.path = "figs/",
  echo = TRUE
)

```

[![Travis-CI Build Status](https://travis-ci.org/mhinsch/rpathsonpaths.svg?branch=master)](https://travis-ci.org/mhinsch/rpathsonpaths)
[![Coverage Status](https://codecov.io/github/mhinsch/rpathsonpaths/coverage.svg?branch=master)](https://codecov.io/github/mhinsch/rpathsonpaths?branch=master)


# Installing the package

Prerequisites:

- C++ boost libraries

- *Rcpp* (>= 0.12.9)

Some of the functionality will also require the *igraph* package. 

*rpathsonpaths* is not on CRAN yet, therefore for now please install the development version of the package from *github*:
```{r install2, eval=FALSE}
devtools::install_github("mhinsch/rpathsonpaths")
```

Note that this requires the package *devtools* installed.


# What does it do?

*rpathsonpaths* provides functions to simulate the spread of a population in a substrate that itself is transported across
a directed graph. The paradigmatic example is the spread of pathogens in food transport networks, but other instances might
include for examples infections in migrating animals.


The most important functions include:

- **`popsnetwork`**: create a network object from an edgelist (describing the transport graph) and a list of external sources.

- **`set_allele_freq`**: set the population composition in source nodes.

- **`popgen_dirichlet`**: simulate spread of genetic material on the network using a Dirichlet distribution to approximate genetic drift.

- **`draw_isolates`**: draw a number of random individuals from a set of nodes using the simulated allele frequencies.

- **`plot`**: plot a *popsnetwork* object (currently uses igraph, but more backends are planned).


# Resources

## Vignettes

An overview of *rpathsonpaths* is provided below in the worked example below.
More detailed tutorials are distributed as vignettes with the package:
```{r, vignettes2, eval = FALSE}
vignette("overview", package="rpathsonpaths")
```


## Websites

The following websites are available:

- The *rpathsonpaths* project on *github*, useful for developers, contributors, and users wanting to post issues, bug reports and feature requests: <br>
[http://github.com/mhinsch/rpathsonpaths](http://github.com/reconhub/rpathsonpaths)


## Getting help online

Bug reports and feature requests should be posted on *github* using the [*issue*](http://github.com/mhinsch/rpathsonpaths/issues) system. 


# A quick overview

The following worked example provides a brief overview of the package's
functionalities. See the [*vignettes section*](#vignettes) for more detailed tutorials.


## Preparing the network


First, we prepare an edge list describing the network, and the initial state of the source nodes:

```{r, prepare}
library(rpathsonpaths)

inp <- c(0L, 0L, 1L, 2L, 3L, 1L, 5L)
outp <- c(1L, 2L, 3L, 3L, 4L, 4L, 2L)
rates <- c(1, 1.5, 0.5, 0.1, 1, 0.1, 0.5)
edgelist <- data.frame(inp, outp, rates)
ext <- data.frame(c(0L, 5L), c(0.5, 0.5))

```

Now we can create a *popsnetwork* object:

```{r, popsnetwork}

netraw <- popsnetwork(edgelist, ext, 0.1)
netraw
plot(netraw)

```


## Injecting initial populations and simulating spread

We set up the initial populations of the source nodes:

```{r, inipop}
freqs <- matrix(c(0.1, 0.4, 0.3, 0.1, 0.2, 0.2, 0.4, 0.3), ncol=4, nrow=2)
netini <- set_allele_freqs(netraw, list(c(0L, 5L), freqs))
netini

```

Now we are ready to simulate spread of genetic material in the network. Note that every call to popgen_dirichlet
will generally produce a different allele frequency distribution.

```{r, dirichlet}
netdir1 <- popgen_dirichlet(netini, 1.7)
netdir1
netdir2 <- popgen_dirichlet(netini, 1.7)
netdir2
```


## Drawing samples

Finally we can draw samples from our simulated population.

```{r, isolates}
samplconf <- data.frame(nodes=c(2L, 4L), N=c(10L, 10L))
draw_isolates(netdir1, samplconf)
draw_isolates(netdir2, samplconf)

```




# Contributors (in alphabetic order):
- [Martin Hinsch](https://github.com/mhinsch)
- [Thibaut Jombart](https://github.com/thibautjombart)

See details of contributions on:
<br>
https://github.com/mhinsch/rpathsonpaths/graphs/contributors


Contributions are welcome via **pull requests**.

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

**Maintainer:** Martin Hinsch (hinsch.martin@gmail.com)
