---
title: "Overview of the rpathsonpaths package"
author: "Martin Hinsch"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width=7, 
  fig.height=5, 
  fig.path="figs-overview/"
)
```


*rpathsonpaths* provides functions to simulate the spread of biological organisms on a substrate that is itself moving through a network structure (e.g. pathogens in a food distribution network or infections in migrating animals). This vignette provides an overview of current features. It largely reproduces the
content of `REAME.md`.



<br>

# Installing the package

*rpathsonpaths* is currently only available on github.

Prerequisites:

- *Rcpp* (>= 0.12.9)

Some of the functionality will also require the *igraph* package. 


```{r install2, eval=FALSE}
devtools::install_github("reconhub/incidence")
```

Note that this requires the package *devtools* to be installed.



<br>

# What does it do?


*rpathsonpaths* provides functions to simulate the spread of a population in a substrate that itself is transported across
a directed graph. The paradigmatic example is the spread of pathogens in food transport networks, but other instances might
include for examples infections in migrating animals.


The most important functions include:

- **`popsnetwork`**: create a network object from an edgelist (describing the transport graph) and a list of external sources.

- **`set_allele_freq`**: set the population composition in source nodes.

- **`popgen_dirichlet`**: simulate spread of genetic material on the network using a Dirichlet distribution to approximate genetic drift.

- **`draw_isolates`**: draw a number of random individuals from a set of nodes using the simulated allele frequencies.

- **`plot`**: plot a *popsnetwork* object (currently uses *igraph*, but more backends are planned).


<br>

# A quick overview

The following worked example provides a brief overview of the package's
functionalities. 


## Preparing the network


First, we prepare an edge list describing the network, and the initial state of the source nodes:

```{r, prepare}
library(rpathsonpaths)

# from nodes
inp <- c(0L, 0L, 1L, 2L, 3L, 1L, 5L)
# to nodes
outp <- c(1L, 2L, 3L, 3L, 4L, 4L, 2L)
# rate of transport, this is optional
rates <- c(1, 1.5, 0.5, 0.1, 1, 0.1, 0.5)
edgelist <- data.frame(inp, outp, rates)
```

We have to provide some external input into the network (otherwise the
simulation will be quite boring). In this case we only give the IDs of the
input nodes and the (absolute) amount of infected material which means 1 will
be assumed for net input rates (infected + uninfected). This is fine since we
intend to run the Dirichlet model so absolute numbers are irrelevant. If we
wanted to run the mechanistic model we would need both sets of numbers.

```{r, prepare2}
# external input (node IDS, infected input)
ext <- data.frame(c(0L, 5L), c(0.5, 0.5))
```

Now we can create a *popsnetwork* object:

```{r, popsnetwork}
# create network using infection rate (within nodes) of 0.1
# this already calculates the amount of infected/unifected for
# all nodes
netraw <- popsnetwork(edgelist, ext, 0.1)
netraw
plot(netraw)
```

We have now set up the network topology and determined the amount of infected
and uninfected material in each node. In the next step we have to set allele
frequencies at the input nodes. Then we can run the simulation to obtain
genetic composisition at the leaf nodes.


## Injecting initial populations and simulating spread

We set up the initial genetic composition of the input nodes. In this case we
assume all root nodes have external input.

```{r, inipop}
# allele frequencies, rows=nodes, columns=alleles
freqs <- matrix(c(0.1, 0.4, 0.3, 0.1, 0.2, 0.2, 0.4, 0.3), ncol=4, nrow=2)
netini <- set_allele_freqs(netraw, list(c(0L, 5L), freqs))
netini
```

Now we are ready to simulate spread of genetic material in the network. Note that every call to popgen_dirichlet
will generally produce a different allele frequency distribution.

```{r, dirichlet}
# simulate with a theta of 1.7
netdir1 <- popgen_dirichlet(netini, 1.7)
netdir1
netdir2 <- popgen_dirichlet(netini, 1.7)
netdir2
```


## Drawing samples

Finally we can draw samples from our simulated population.

```{r, isolates}
# 10 samples from nodes 2 and 4
samplconf <- data.frame(c(2L, 4L), c(10L, 10L))
draw_isolates(netdir1, samplconf)
draw_isolates(netdir2, samplconf)

```

# An example using real data

## Obtain and clean up data

Let's load the data:

```{r}
library(outbreaks)
data(s_enteritidis_pt59)
# outbreaks stores node lists as string vectors...
raw_graph = s_enteritidis_pt59$graph
# ...but putting them into a dataframe will convert them to factors
sent_graph = data.frame(from=raw_graph[[1]], to=raw_graph[[2]])
```

Just in case we do some sanity checks.

```{r}
library(rpathsonpaths)
# check for separate subnetworks
colour_network(sent_graph)
# check for cycles
cycles(sent_graph)
```

Ok, we do have one contiguous network. Unfortunately, however, there's a cycle. 
This will make rpathsonpaths very unhappy so we have to get rid of it.

```{r}
# output cycles
cycles(sent_graph, TRUE)
sent_graph2 <- subset(sent_graph, !(from=="13dc78" & to=="0b6e5a"))
# success?
cycles(sent_graph2)
```

## Create network

Ok, this is solved. Now let's see how many root nodes we have in the network.

```{r}
sg_sources <- sources(sent_graph2)
sg_sources
```

In order to be able to construct a popsnetwork object we have to provide at least the rate of
input of infected material for each source.

```{r}
# this is the lazy way
sg_ext <- data.frame(sg_sources, rep(0.5, length(sg_sources)))
sg_ext
```

Now we can create the network.

```{r}
# some mild infection within nodes
sg_net <- popsnetwork(sent_graph2, sg_ext, 0.1)
plot(sg_net)
```

## Run simulation

The only thing left to do in order to run the simulation is to provide initial allele frequency distributions for the source nodes.

```{r}
# it's just an example so let's make things easy
sg_freqs <- matrix(rep(c(0.25, 0.25, 0.5), each=7), nrow=7)
# we can provide allele frequencies when calling the simulation
sg_net_dir <- popgen_dirichlet(sg_net, 1.3, list(sg_sources, sg_freqs))
sg_net_dir
```

Now we are ready to draw lots of isolates.

```{r}
# leaf nodes
sg_sinks <- sinks(sent_graph2)
# 20 samples from each sink node
draw_isolates(sg_net_dir, data.frame(sg_sinks, rep(20, length(sg_sinks))))
```

# Multiple runs and analysis

A single simulation is not particularly interesting. In this section we will
therefore run a whole set of simulations and do some basic analysis on the
results. 

As a simple example we will try to find out whether topological distance between
nodes on the network results in a corresponding genetic distance when looking at
the genetic composition of the nodes' pathogen population.

For the sake of simplicity we will use a simple binary tree for this example.
Note that for maximum flexibility `perfect_binary` creates an edge list, *not*
a `popsnetwork` object.

```{r}
# generate network
tree <- perfect_binary(4L)
```

We measure genetic distance between two samples as the Hamming distance of
their genomes and the genetic distance between two nodes as the genetic
distance of two random isolates drawn from these nodes. 

By default rpathsonpaths only simulates single loci. This would result in all genetic
distance measurements to be either `0` or `1`, which is not very useful for statistical
analysis. It is quite simple, however, to emulate multiple loci - simply run the same 
scenario multiple times and collate the results (note that this assumes that there is
no linkage disequilibrium).

We first have to generate initial allele frequencies for each locus.

```{r}
# generate a list of random allele frequencies
# these are the initial allele frequency distributions at 10 "loci"
all_freqs <- replicate(10, list(0, gtools::rdirichlet(1, c(1, 1))), simplify=FALSE)
```

`run_popsnet` is a convenience frontend for `popsnetwork`, `set_allele_freqs`
and the `popgen_*` simulator functions. It makes it straightforward to run a
large set of simulations on the same network topology. Note that, since the
function calls `popsnetwork`'s contructor itself, the network has to be
provided in edgelist format. The result of a call to `run_popsnet` is a simple
list containing the raw network object (as produced by the contructor) and a
vector of network objects containing the simulation data.

```{r}
# now we can run the simulation
# since all_freqs is a list the simulation will be run n (in this case 1) times
# per element
results <- run_popsnet(tree, ini_input=1, ini_infd=0.5, 
					   ini_freqs=all_freqs, n=1L, theta=0.1,
					   # we use the Dirichlet model
					   spread_model="fluid", drift_model="dirichlet")
```

Now we have to do something with the results. We will first obtain topological
distances using rpathsonpaths' `path_distances` function. Since in our simple
scenario infection cases equal leaf nodes of the network we keep only the distances
between all leaf nodes.

```{r}
# $raw contains the pre-simulation net
raw <- results$raw
# we need those since we are only interested in leaf nodes
snks <- as.character(sinks(edge_list(raw)))
# get topological distances of all pairs of leaf nodes
path_dists <- path_distances(raw)[snks, snks]
```

Next we determine genetic distances. The builtin function `distances_sample`
does most of the heavy lifting for us. It draws a random isolate from each node
and calculates Hamming distance for each pair of nodes. As before we throw away
everything but the leaf nodes. This will give us a matrix of 1s and 0s for each
of the simulations, which - as we said before - correspond to one locus each.
We simply sum all results to obtain genetic distance over the whole genome.

```{r}
# the actual results
res <- results$result
# this runs distances_sample on each of the networks in the results and sums
# up the results
gene_dists <- Reduce('+', lapply(res, function(x) distances_sample(x)[snks, snks]))
```

Now we can plot genetic versus topological distance and the resulting correlation. As we
can see there is a clear correlation between topological and genetic distance. 

```{r}
plot(x=jitter(as.vector(path_dists), 0.1), y=jitter(as.vector(gene_dists), 0.02), xlab="topological", ylab="genetic")
abline(lm(as.vector(gene_dists)~as.vector(path_dists)), col="red")
```
